---
title: "Exploratory Analysis"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(knitr)
library(tidyverse)
library(dplyr)
library(readxl)
library(ggplot2)
```

## Project title

Global Population Dynamics: The Role of Birth, Death, and Life Expectancy

### Authors

Aaron Chen (chen716@uw.edu)
Sam Junxuan Chen (junxuan@uw.edu)
David Sung Ha Hyun (davyhyun@uw.edu)

### Date

Spring 2024
05/14/2024

### Abstract

"Our main question is understand the relationship between demographic factors such as life expectancy, birth rate, and death rate with population size. This question is important because it can provide insights into the impact of population dynamics on human well-being and societal trends. To address the question, we will analyze the correlations between these variables like life expectancy and fertility rate and their relationships."

### Keywords

Social Science, Global Population, Population dynamics

### Introduction  

Our project aims to unravel the complexity of population dynamics by investigating the relationship between different demographic factors, such as life expectancy, birth rate, and death rate, and overall population size. These factors are critical to understanding the broader consequences of demographic shifts for society structures and individual well-being. Our project raises several research questions, including:
- How does life expectancy correlate with changes in population size across different regions?
- What impact do variations in birth and death rates have on the population size of different countries?
- Which factors, life expectancy, birth rate, death rate, or fertility rate contribute most to population?


The motivations behind these research questions are rooted in understanding how demographic shifts can influence a wide range of societal and economic outcomes. Through these questions we can assess how increases in life expectancy influence regional population structures and the associated economic and social planning needs; we can analyze how changes in birth and death rates drive population growth or decline, informing resource allocation, infrastructure development, and economic strategies; and we can identify the most impactful demographic factors on population dynamics to determine the effectiveness of different societal factors such as healthcare, education, and economic planning.


These questions are important because they address fundamental issues about human growth and social stability. Understanding these dynamics enables politicians, researchers, and public health authorities to better predict future trends and develop solutions to promote health, economic stability, and quality of life. 

### Related Work  

Our research examines the complex relationships between demographic characteristics such as life expectancy, birth rate, and death rate, as well as their effects on population size. This issue matters because it addresses fundamental aspects of economic, social, and health-related processes across many geographical and socioeconomic contexts. Understanding these relationships contributes in addressing concerns such as aging populations, workforce dynamics, healthcare delivery, and social security systems.

[Walker, B. H., & Brown, D. C."Trends in lifespan variation across the spectrum of rural and urban places in the United States, 1990–2017."](https://www.ncbi.nlm.nih.gov/pmc/articles/PMC9434220/):  This study highlights the growing disparities in mortality rates and life expectancy between urban and rural areas in the U.S., emphasizing the importance of geographical differences in demographic research. 

[Roser, M. (2019). "Demographic transition: Why is rapid population growth a temporary phenomenon?"](https://ourworldindata.org/demographic-transition#:~:text=Death%20rates%20fall%20first%2C%20then,a%20slowdown%20of%20population%20growth.&text=Population%20growth%20is%20determined%20by,have%20fallen%20across%20the%20world.): Max Roser explains the concept of demographic transition through historical and contemporary examples, explaining how and why populations grow and eventually stabilize. This work is crucial for understanding the temporal dynamics of population changes and the roles that mortality and fertility rates play. 

[“Main Factors Driving Population Growth”](https://www.pewresearch.org/religion/2015/04/02/main-factors-driving-population-growth/): This source delves into the primary drivers of population growth, such as fertility rates, mortality rates, age profiles, migration, and religious switching. It provides a comprehensive overview of how these factors individually and collectively influence population trends and offers projections that consider different scenarios, thereby contributing to a nuanced understanding of demographic changes.

These sources serve as a foundation for comprehending the complexities of demographic change and their worldwide repercussions. They also emphasize the importance of studying population dynamics in order to better predict future demographic transitions and develop suitable policy responses.

### The Dataset

The data was found on Kaggle. Here is the link to the data source: [Global Population Trends(2016-2022)](https://www.kaggle.com/datasets/alitaqi000/global-population-trends2016-2022). The data was collected by Syed Ali Taqi. The data compiled data from Macrotrends, [Macrotrends Offical Website](https://www.macrotrends.net/). Macrontends is a company that provides a standard stock screener and their data collection methodology is to take information from government records and integrate that data with economic data, financial market data, company expert judgment and data. The data using in the project are cited by Syed Ali Taqi from Macrontends' data integration of Population. There are 1074 observations (rows) in the dataset.There are 12 features (columns) in the dataset. In conducting the analysis, conclusions drawn from the data may have ethical implications, especially if the conclusions contribute to the perpetuation of stereotypes or stigmatization of certain groups of the population. It is important to consider the impact that the data may have on different communities and whether the use of the data is consistent with principles of social justice and equity. A limiting factor may be the accuracy and reliability of the data, as demographic data collection methods may differ across countries and regions, leading to inconsistencies and potential bias. In addition, missing or incomplete data may affect the comprehensiveness of the analysis. In addition, demographic indicators may not fully capture the complexity of population dynamics and certain population groups like minors may be underrepresented in the data.

### Implications

If the research in our project succeeds in answering the questions posed, we believe that technicians, designers, and policymakers will be greatly affected. First, for technicians, understanding the impact of demographic factors on population dynamics can help them build models for prediction in fields such as public health, urban planning, and resource management. For designers, they can use this research to create more adaptable and sustainable living environments to meet changing demographic characteristics, such as the aging of the population that many countries in the world are now facing. Or the problem of rapid urbanization. For policymakers, they will have very comprehensive data to analyze issues such as education, health care and pensions, and can help them solve social phenomena such as overpopulation, underpopulation or an aging workforce. This multidisciplinary approach ensures that social infrastructure develops in step with demographic trends, promoting stability and prosperity.

### Limitations & Challenges
In our opinion, a significant limitation of our project is the challenge of data heterogeneity and standardization. Demographic data may be collected using different methods in different countries and regions. This can lead to inconsistencies in comparative analyses, for example, the indicators measuring birth and death rates may differ, or the definition of “urban” versus “rural” may differ. Differences in these indicators will affect the accuracy of the analysis. In addition, demographic characteristics change rapidly due to factors such as immigration, policy changes, or major economic shifts. This means that not only can the data we use be inconsistent, but it can also quickly become outdated. Therefore, any conclusions drawn must be considered within the context of limitations and are subject to the possibility of bias and error. The project will require rigorous data validation methods and adaptive approaches to accommodate and reconcile these differences to ensure findings are reliable and reflect true population dynamics. 

### Summary Information

Write a summary paragraph of findings that includes the 5 values calculated from your summary information R script

### Table

Include a table of aggregate information

Describe why you included the table and what information it reveals

### Chart 1

Include a chart

Describe why you chose this chart and what information it reveals

Here's an example of how to run an R script inside an RMarkdown file:

``````{r , echo = FALSE, code = readLines("chart1_example.R")}
```

### Chart 2: Top 5 and Bottom 5 Countries by Average Life Expectancy (2018-2021)

Include a chart

Describe why you chose this chart and what information it reveals
```{r chart2,  echo = FALSE}
source("scripts/Sam_Avg_LE.R")
plot(chart2)
```
This chart shows the average life expectancy of different countries across the globe in 2018-2021, that is, how many years the population of that country will live on average. It shows the 5 countries with the highest life expectancy and the 5 countries with the lowest life expectancy.

### Purpose
The purpose of this chart is to examine the variation in average life expectancy across different countries from 2018 to 2021. By identifying the top five countries with the highest life expectancy and the bottom five countries with the lowest life expectancy, we aim to gain insights into global demographic trends. 

## Insights
- The five countries with the lowest average life expectancy are all located on the African continent, and none of them has a life expectancy above 60.
 -These lowest ranked countries are expected to be deficient in economic stability, quality of life, local government strategic decisions, resource allocation, and social planning.
- The five countries with the highest average life expectancy are all outside the African continent, with three countries from Europe and two from Asia.
  - They are all developed countries with good education, health and social systems. However, they have the best resources, which means that the distribution of resources and social planning have a significant impact on the expected age.

### Chart 3

Include a chart

Describe why you chose this chart and what information it reveals
